---
title: "plotly"
description: "argh"
author: Simon Coulombe
date: 2024-09-24
categories: [r, plotly]
lang: fr
---

```{r setup, echo = FALSE}
library(ggplot2)
library(plotly)
library(dplyr)
library(gapminder)
library(listviewer)
pal <- ggokabeito::palette_okabe_ito()
df <- gapminder::gapminder %>% 
  filter(country %in% unique(gapminder::gapminder$country)[1:8]) %>% 
  mutate(date = as.Date(paste0(year, "-01-01")),
         lifeExp = round(lifeExp, 2)) %>% 
  filter(year >= 1990) %>% 
  group_by(year) %>% 
  mutate(country = factor(country, levels = unique(.$country)),
         pop_percent =  pop / sum(pop)) %>%
  ungroup()

df_long <- df %>% 
  select(country, date, lifeExp, pop_percent) %>% 
  tidyr::pivot_longer(
    cols = all_of(c("lifeExp", "pop_percent")),
    names_to = "variable",
    values_to = "value"
  )
```
# use plotly::schema()


schema allows you to browse available properties
```{r}
library(listviewer)
plotly::schema()
```


# Building a plotly   



the absolute minimum

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines")
```

i like this hovermode   

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified") 
```

adding titles 

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = "this is my title"))
```

adding a subtitle inside the title

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>')
    )
  )
```

left-align the title    

```{r}

plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left"
    )
  )

```



add some margin  above the title   , remove  some margin to the left of the y axis and below the x axis


```{r}

plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left"
    )
  ) %>%
  plotly::layout(
    margin = list(
      #l = 50, # less than default
      # r = 40,
      #b = 40, 
      t = 50#,
      #pad = 40
      
    )
  )


```


add a title to the legend

```{r}

plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left"
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country")))
```


use the x_axis title as a caption.

I could also use an annotation, but I have having the guess how much margin I have to add and where to position it.
https://stackoverflow.com/questions/45103559/plotly-adding-a-source-or-caption-to-a-chart


```{r}

plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left"
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country"))) %>%
  plotly::layout(
    xaxis = list(
      title = list(text = "<sup>source: gapminder</sup>")
    )
  )
```

adjust the title size to fit ggplot theme.   

```{r}

plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left",
                 font = list(size =ggplot2::theme_get()$text$size  * 1.4)
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country",
                                            font = list(size = ggplot2::theme_get()$text$size)),
                               font = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
  )
  ) %>%
  plotly::layout(
    xaxis = list(
      title = list(text = "<sup>source: gapminder</sup>",
                   font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  ) %>% 
  plotly::layout(
    yaxis = list(
      title = list(text = "lifeExp", font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  )  
```

pick the color palette.
note, when passing a vector of colors, you have to give the exact number of colors, otherwise it will interpolate.  

here is my palette:
```{r}
scales::show_col(pal)
```



here is what happens if I don't have the correct number of colors in the palette (9, colors, 8 countries).. what happens from "yellow" onward?

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                colors = pal,
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('Colors randomly interpolated because I provided 9 colors and 8 countries',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left",
                 font = list(size =ggplot2::theme_get()$text$size  * 1.4)
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country",
                                            font = list(size = ggplot2::theme_get()$text$size)),
                               font = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
  )
  ) %>%
  plotly::layout(
    xaxis = list(
      title = list(text = "<sup>source: gapminder</sup>",
                   font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  ) %>% 
  plotly::layout(
    yaxis = list(
      title = list(text = "lifeExp", font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  )  
```




here is the proper way (pal[1:8])

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ lifeExp, 
                color =  ~country, 
                colors = pal[1:8],
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left",
                 font = list(size =ggplot2::theme_get()$text$size  * 1.4)
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country",
                                            font = list(size = ggplot2::theme_get()$text$size)),
                               font = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
  )
  ) %>%
  plotly::layout(
    xaxis = list(
      title = list(text = "<sup>source: gapminder</sup>",
                   font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  ) %>% 
  plotly::layout(
    yaxis = list(
      title = list(text = "lifeExp", font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
    )
  )  
```

format y axis as percent, format x axis as date
note, I used %-d  instead of %d to remove left-padding with 0  ( I want Jan 1,  not Jan 01)

```{r}
plotly::plot_ly(df, x = ~ date, 
                y = ~ pop_percent, 
                color =  ~country, 
                colors = pal[1:8],
                type = "scatter", 
                mode = "markers+lines") %>%
  plotly::layout(hovermode="x unified")  %>%
  plotly::layout(
    title = list(text = paste0('this is my title',
                               '<br><sup>This is my subtitle','</sup>'),
                 x = 0.02,
                 xanchor = "left",
                 font = list(size =ggplot2::theme_get()$text$size  * 1.4)
    )
  ) %>%
  plotly::layout(margin = list(t = 50)) %>% 
  plotly::layout(legend = list(title = list(text = "country",
                                            font = list(size = ggplot2::theme_get()$text$size)),
                               font = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size ))
  )
  ) %>%
  plotly::layout(
    xaxis = list(
      title = list(text = "<sup>source: gapminder</sup>",
                   font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size )),
      tickformat = "%b %-d<br>%Y"
    )
  ) %>% 
  plotly::layout(
    yaxis = list(
      title = list(text = "pop_percent", font = list(size = ggplot2::theme_get()$text$size)),
      tickfont = list(size = ggplot2::theme_get()$text$size * as.numeric(ggplot2::theme_get()$axis.text$size )),
      tickformat = ".0%"
    )
  )  
```




# A single_plotly_timeseries function  

```{r}
single_plotly_timeseries <-  function(df,  x_var, y_var, color_var = NULL, caption = NULL, title = NULL, subtitle = NULL,text_size = 15, tickformat = NULL, showlegend = TRUE,  ...) {
  
  
  if (!is.null(subtitle)){ complete_title <- paste0(title, "<br><sup>", subtitle, "</sup>")}
  else{complete_title <- title}
  
  # if(!is.null(title)){ top_margin <- pmax(50, text_size+2)
  # } else {top_margin <- pmax(15, text_size+2) }
  
  top_margin <- pmax(15, text_size+2) +  as.numeric(!is.null(title))*2*text_size + as.numeric(!is.null(subtitle)) *2 *text_size
  max_abs_y_var <- max(abs(df[y_var]), na.rm = TRUE)
  
  plotly::plot_ly(df, x = ~.data[[x_var]], y = ~.data[[y_var]], ...) %>%  
    { if(!is.null(color_var)){ # add color var if required
      plotly::add_trace(.,
                        type = "scatter",
                        mode = "markers+lines+text",
                        color =  ~.data[[color_var]],
                        showlegend = showlegend
      ) %>%
        plotly::layout(
          legend = list( # update legend fond and size
            font = list(size = text_size * 0.8),
            title = list(
              text = color_var,
              font = list(size = text_size)
            )
          )
        )
    } else{
      plotly::add_trace(.,
                        type = "scatter",
                        mode = "markers+lines+text"
      )
    }
    }  %>% 
    plotly::layout(hovermode="x unified")  %>%
    plotly::layout(
      title = list(text = complete_title,
                   x = 0.02,
                   xanchor = "left",
                   font = list(size =text_size  * 1.4)
      )
    ) %>%
    plotly::layout(margin = list(t = top_margin)) %>%
    plotly::layout(
      xaxis = list(
        title = list(text = paste0("<sup>", caption, "</sup>") ,
                     font = list(size = text_size)),
        tickfont = list(size = text_size * 0.8),
        tickformat = "%b %-d<br>%Y"
      )
    ) %>%
    plotly::layout(
      yaxis = list(
        title = list(text = y_var, font = list(size = text_size)),
        tickfont = list(size = text_size * 0.8)#,
        #tickformat = ".0%"
      )
    ) %>% 
    { if(!is.null(tickformat)){ # if tickformat is set, use it
      plotly::layout(.,  yaxis = list(tickformat = tickformat))
    } else if (max_abs_y_var < 1){ # if no forced tickformat and maximum absolute value <1, then use percent
      plotly::layout(.,  yaxis = list(tickformat = ".0%"))
    } else {  # just use the default tick format
      .}
    } 
}
```

function demo:
```{r}
single_plotly_timeseries(df, "date", "lifeExp", "country", title = "Clever title", subtitle = "cunning subtitle", caption = "source: gapminder")
```

function demo, auto set yaxis to format : 
```{r}
single_plotly_timeseries(df, "date", "pop_percent", "country", title = "Clever title", subtitle = "cunning subtitle", caption = "source: gapminder")
```

function demo, use ... to set height
```{r}
single_plotly_timeseries(df, "date", "pop_percent", "country", title = "Clever title", subtitle = "cunning subtitle", caption = "source: gapminder", height = 800)
```



# subplots  

```{r}
p1 <- plot_ly(economics, x = ~date, y = ~unemploy) %>% 
  add_lines(name = "unemploy")
p2 <- plot_ly(economics, x = ~date, y = ~uempmed) %>% 
  add_lines(name = "uempmed")
subplot(p1, p2)
```
## subplot with colors  
adding colors naively lets to having each country twice in the legend : 

```{r}
p1 <- plot_ly(df, x = ~date, y = ~lifeExp, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter")
p2 <- plot_ly(df, x = ~date, y = ~pop_percent, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter") 
subplot(p1, p2)
```

I must specify showlegend= FALSE in all the subplots after the first one: 
```{r}
p1 <- plot_ly(df, x = ~date, y = ~lifeExp, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter")
p2 <- plot_ly(df, x = ~date, y = ~pop_percent, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter", showlegend = FALSE)  # only add color legend in first plot
subplot(p1, p2)
```


I can specify nrows and shareX. titleY is also fun, to keep the y-axis title
```{r}
p1 <- plot_ly(df, x = ~date, y = ~lifeExp, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter")
p2 <- plot_ly(df, x = ~date, y = ~pop_percent, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter", showlegend = FALSE)  # only add color legend in 1 place
subplot(p1, p2, nrows = 2, shareX= TRUE, titleY = TRUE)
```


test with single_plotly_timeseries
note: they must have the same title / caption otherwise it doesnt get displayed?
```{r}

p1 <- single_plotly_timeseries(df, "date", "pop_percent", "country", title = "Clever title", subtitle = "cunning subtitle", caption = "source: gapminder", showlegend = TRUE)

p2 <- single_plotly_timeseries(df, "date", "lifeExp", "country", title = "Clever title", subtitle = "cunning subtitle", caption = "source: gapminder", showlegend = FALSE) 

subplot(p1, p2, nrows = 2, shareX= TRUE, titleY = TRUE)
```




# facets    


a function for paneling found in Carson's book  https://plotly-r.com/arranging-views

```{r}
panel <- .  %>% 
  plot_ly(x = ~date, y = ~value) %>%
  add_lines() %>%
  add_annotations(
    text = ~unique(variable),
    x = 0.5,
    y = 1,
    yref = "paper",
    xref = "paper",
    yanchor = "bottom",
    showarrow = FALSE,
    font = list(size = 15)
  ) %>%
  layout(
    showlegend = FALSE,
    shapes = list(
      type = "rect",
      x0 = 0,
      x1 = 1,
      xref = "paper",
      y0 = 0, 
      y1 = 16,
      yanchor = 1,
      yref = "paper",
      ysizemode = "pixel",
      fillcolor = toRGB("gray80"),
      line = list(color = "transparent")
    )
  )


economics_long %>%
  group_by(variable) %>%
  do(p = panel(.)) %>%
  subplot(nrows = NROW(.), shareX = TRUE)
```



naively adding annotations to my colored plot 


```{r}
p1 <- plot_ly(df, x = ~date, y = ~lifeExp, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter") %>% 
  add_annotations(
    text = "lifeExp",
    x = 0.5,
    y = 1,
    yref = "paper",
    xref = "paper",
    yanchor = "bottom",
    showarrow = FALSE,
    font = list(size = 15)
  ) %>%
  layout(
    shapes = list(
      type = "rect",
      x0 = 0,
      x1 = 1,
      xref = "paper",
      y0 = 0,
      y1 = 16,
      yanchor = 1,
      yref = "paper",
      ysizemode = "pixel",
      fillcolor = toRGB("gray80"),
      line = list(color = "transparent")
    )
  )
p2 <- plot_ly(df, x = ~date, y = ~pop_percent, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter", showlegend = FALSE) %>%    # only add color legend in 1 place
  add_annotations(
    text = "pop_percent",
    x = 0.5,
    y = 1,
    yref = "paper",
    xref = "paper",
    yanchor = "bottom",
    showarrow = FALSE,
    font = list(size = 15)
  ) %>%
  layout(
    shapes = list(
      type = "rect",
      x0 = 0,
      x1 = 1,
      xref = "paper",
      y0 = 0,
      y1 = 16,
      yanchor = 1,
      yref = "paper",
      ysizemode = "pixel",
      fillcolor = toRGB("gray80"),
      line = list(color = "transparent")
    )
  )
subplot(p1, p2, nrows = 2, shareX= TRUE)


```

```{r}
add_facet_title <- function(plotly, facet_title, text_size  = 15){
  plotly %>% 
    add_annotations(
      text = facet_title,
      x = 0.5,
      y = 1,
      yref = "paper",
      xref = "paper",
      yanchor = "bottom",
      showarrow = FALSE,
      font = list(size = text_size)
    ) %>%
    layout(
      shapes = list(
        type = "rect",
        x0 = 0,
        x1 = 1,
        xref = "paper",
        y0 = 0,
        y1 = (text_size+1),
        yanchor = 1,
        yref = "paper",
        ysizemode = "pixel",
        fillcolor = toRGB("gray80"),
        line = list(color = "transparent")
      )
    )
}
```
test de ma fonction à panel title  
```{r}
plot_ly(df, x = ~date, y = ~lifeExp, color = ~ country) %>% 
  add_trace(mode = "markers+lines", type = "scatter") %>%
  add_facet_title("lifeExp") 
```
```{r}
df_plots <- df_long %>% 
  dplyr::nest_by(variable) %>% 
  ungroup() %>% 
  dplyr::mutate(
    first_row  = row_number() == 1) %>%
  rowwise() %>% 
  dplyr::mutate(
    plot = list(
      single_plotly_timeseries(
        data, "date", "value", "country", showlegend = first_row) %>%
        add_facet_title(variable)  %>%
        layout(yaxis= list(title = ""))))

subplot(df_plots$plot, nrows= nrow(df_plots), shareX = TRUE)
```

let's create a function that wraps both

```{r}
plotly_timeseries <- function(df,  x_var, y_var, color_var = NULL, facet_var = NULL, caption = NULL, title = NULL, subtitle = NULL,text_size = 15, tickformat = NULL, showlegend = TRUE,  ...) {
  
  
  if(!is.null(facet_var)){
  
  
  df_plots <- df %>% 
    dplyr::nest_by(.data[[facet_var]]) %>% 
    ungroup() %>% 
    dplyr::mutate(
      first_row  = row_number() == 1) %>%
    rowwise() %>% 
    dplyr::mutate(
      plot = list(
        single_plotly_timeseries(
          data, x_var =  x_var, y_var = y_var, color_var = color_var, 
          caption = caption, title = title, subtitle = subtitle, text_size= text_size, tickformat = tickformat,
          showlegend = as.logical(min(showlegend, first_row)), ...) %>%
          add_facet_title(.data[[facet_var]], text_size)  %>%
          layout(yaxis= list(title = ""))
      )
    )
  
  
  subplot(df_plots$plot, nrows= nrow(df_plots), shareX = TRUE)
  }else {
    single_plotly_timeseries(
          df, x_var =  x_var, y_var = y_var, color_var = color_var, showlegend = showlegend,
          caption = caption, title = title, subtitle = subtitle, text_size= text_size, tickformat = tickformat, ...)


  }
  
  
  
}
```


plotly_timeseries with a facet:
```{r}
plotly_timeseries(df_long, 
                   x_var = "date",
                   y_var =  "value", 
                   color_var = "country",
                   facet_var = "variable", 
                   title = "My title", 
                   subtitle = "my subtitle", 
                   caption = "source: gapminder")  
```


plotly_timeseries without  facet:
```{r}
plotly_timeseries(df, x_var = "date",
                   y_var =  "lifeExp", 
                   color_var = "country", 
                   showlegend = TRUE, 
                   title = "My title", 
                   subtitle = "my subtitle", 
                   caption = "source: gapminder")
```

